Viscous Erosion in Porous Media

### 2018-01-16 docs/methods/erosion_methods.tex
small changes to intro

Because of the scales involved with groundwater flow, we will restrict
our attention to Stokesian dynamics.  For groundwater flow, Darcy's
equation are often used, but this assumes a homogeneous geometry which
is not always a fair assumption.  Instead, we will use the
incompressible Stokes equations to represent the flow, and the
individual pores will be be resolved, resulting in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited as it can resolve complex geometries and the non-negligible
interactions between bodies are resolved, all the while high-order
accuracy and computational efficiency are achieved. To link the body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  To further visualize the effect
of erosion, we compute the vorticity of the flow
(Figure~\ref{fig:50bodies}) which is equal to the shear stress in the
Stokesian limit.  Finally, to evolve the geometry, a second-order time
stepping method that guarantees that the mesh does not become distorted
will be employed.

\paragraph{Contributions} The present study will develop numerical
methods to simulate viscous erosion of multiply-connected
two-dimensional geometries.  By assuming the flow is viscous, we are
assuming the flow is very slow and in small length scales, as is the
case in groundwater flow.  Moreover, by allowing for complex
multiply-connected domains, the effects due to interacting bodies on the
erosion are captured.  We start by authenticating our method on a single
circular body placed symmetrically in a pipe flow.  By artificially
fixing the area of the body as a function of time, we are able to
characterize the long-time behavior of the shape and the resulting flow.
Then, we compute the vorticity of erosion in multiply-connected
geometries to better characterize erosion in porous media.

@@

\todo[inline]{DUMP OF TEXT}
This choice is based on experimental work of Ristroph et
al.~\cite{ris-moo-chi-she-zha2012}, but other models for erosion are
possible.  For example, some models use an erosion rate that is
proportional to a power of the magnitude of the shear
stress~\cite{par-izu2000}, while others introduced a shear stress below
which erosion does not occur~\cite{lag2000}.
\todo[inline]{DUMP OF TEXT}

\paragraph{Limitations} There are two main limitations of the current
work.  First, the flow is assumed to be two-dimensional.  However, by
assuming that the flow is two-dimensional, we are able to simulate the
erosion of more bodies and with higher resolution than would be
available in three dimensions.  Moreover, Several flows can be described
to good accuracy under a two-dimensional approximation, certain
phenomena occur only in three dimensions.  The boundary integral
equation formulation for the fluid solver exists, and the main
challenges of a BIE solver have been addressed by other groups including
surface representation~\cite{yin-bir-zor2006, vee-rah-bir-zor2011},
near-singular integration~\cite{yin-bir-zor2006, kli-tor2016b}, and fast
summation methods~\cite{yin-bir-zor2004, ros-ols2015}.  The more
significant challenge of extending our work to three dimensions is
guaranteeing that the mesh does not become distorted, which can be done,
for example, by using an extension of the two-dimensional {\thL}
formulation~\cite{amb-sie-tlu2013}.

@@

\paragraph{Related work}
Simulating eroding bodies has been studied analytically, experimentally,
and numerically.  Because of the scales involves with geophysical flows,
most of these works consider high Reynolds number flows.  For instance,
the experimental work of Moore et al.~\cite{moo-ris-chi-zha-she2013}
considers on the order of $10^4$.  For low Reynolds number flows,
Mitchell and Spagnolie~\cite{mit-spa2016} simulate the erosion of a
mobile body in three dimensions, but they only consider single body.

@@

\paragraph{Outline of the paper}
In Section~\ref{s:formulation}, we describe the governing equations for
both the fluid and describe the erosion model using a \thL~framework.
In Section~\ref{s:method}, the numerical methods in both space and time
are described.  Section~\ref{s:results} gives several numerical
examples.  Finally, concluding remarks are made in
Section~\ref{s:conclusions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formulation\label{s:formulation}} 
We start by defining the main variables used to model erosion.  We only
consider flows that are confined by a solid wall $\Gamma$ that encloses
$M$ eroding bodies.  The bodies are denoted as $\gamma_\ell$,
$\ell=1,\ldots,M$, and we write $\gamma = \gamma_1 \cup \cdots \cup
\gamma_M$.  Neglecting inertial forces, the dynamics of the fluid is
fully characterized by the position of the bodies $\xx_\ell(s,t) \in
\gamma_\ell$, where $s$ is the arclength and $t$ is time.  Given
$\xx_\ell(s,t)$, $\ell=1,\ldots,M$, derived variables include the fluid
velocity $\uu$, the pressure $p$, and the shear stress $\tau$.   On the
bounding wall, we prescribed a velocity $\UU(\xx,t)$.  Then, the
governing equations are
\begin{equation}
\label{eqn:erosionModel}
\begin{split}
  \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap &&\mbox{conservation
of momentum}\\
\grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap
&&\mbox{conservation of mass} \\
\uu = 0, &\hspace{20pt} \xx \in \gamma, \gap &&\mbox{no slip on the
bodies} \\
\uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap &&\mbox{outer wall
velocity} \\
V_\nn = \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
&&\mbox{erosion model},
\end{split}
\end{equation}
where $V_\nn$ is the normal velocity of the interface and $\mu$ is the
fluid viscosity.  The outer wall is formed by rounding off the corners
of $[-3,3] \times [-1,1]$, and we enforce the Hagen-Poiseuille flow
\begin{align*}
  \uu(\xx) = (1-y^2,0)
\end{align*}
on $\Gamma$.  The boundary of the fluid domain is $\bd\Omega = \gamma_1
\cup \cdots \cup \gamma_M \cup \Gamma$.  To minimize the boundary
effects of the inflow and outflow, we only place bodies in the center
third, $[-1,1] \times [-1,1]$, of the fluid domain.
\begin{figure}[htpb]
  \centering
  \input{figs/schematic.tikz}
  \caption{\label{fig:schematic} A schematic for the governing
    equations.  A no-slip boundary condition is imposed on each body
    $\gamma_\ell$ whose unit normal points outward relative to the
    geometry.  On the outer geometry $\Gamma$, a pipe flow is imposed.
    The bodies are eroded at a rate that is proportional to the
    magnitude of the shear stress applied by the Stokesian fluid.  The
    bodies are constrained to the middle third of the channel that is
    located between the dashed lines.}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods\label{s:method}} 
There are three main numerical methods that need to be developed.  In
Section~\ref{sec:BIE}, we describe how the BIE~\eqref{eqn:completed_BIE}
is solved.  Once the density function is found,
Section~\ref{sec:shearStress} describes how the shear
stress~\eqref{eqn:shearStressLP} is computed.  Then, in
Section~\ref{sec:timeStepping}, a time stepping method for evolving the
interface according to equations~\eqref{eqn:Levo1}
and~\eqref{eqn:thetaevo1} are described.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Post Processing Quantities of Interest\label{s:qoi}}

To study characteristics of the flow, the vorticity
(Section~\ref{sec:vorticity}), pressure (Section~\ref{sec:pressure}),
and drag (Section~\ref{sec:drag}) will be computed.  Once
equation~\eqref{eqn:completed_DLP} is solved for the density function,
Stokeslets, and rotlets, all of these quantities can be computed.  For
target points in the fluid domain, the resulting layer potentials
involve integrands that are periodic, so the trapezoid rule guarantees
spectral accuracy since the integrand is smooth~\cite{tre-wei2014}.
However, if the target point is close to a boundary, a nearly-singular
integrand must be integrated, and this requires upsampling that is
described in Section~\ref{sec:NSI}.  When the target points is on the
boundary, the integrands involve singularities that are evaluated by
combining singularity subtraction and odd-even
integration~\cite{sid-isr1988}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results\label{s:results}}

With the description of the methods complete, we now present numerical results on bodies eroding in Stokes flow. We first discuss the case of a single eroding body, as certain features can be predicted analytically and thus used for validation. We then discuss the erosion of multiple bodies and highlight important differences that arise as compared to the single-body case.

>>>>

Because of the scales involved with groundwater flow, we will restrict
our attention to Stokesian dynamics.  For groundwater flow, Darcy's
equation are often used, but this assumes a homogeneous geometry which
is not always a fair assumption.  Instead, we will use the
incompressible Stokes equations to represent the flow, and the
individual pores will be be resolved, resulting in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited as it can resolve complex geometries and the non-negligible
interactions between bodies are resolved, all the while high-order
accuracy and computational efficiency are achieved. To link the body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}. Other models for erosion include
ones that use an erosion rate that is proportional to a power of the
magnitude of the shear stress~\cite{par-izu2000}, or ones that use a
critical shear stress below which erosion does not occur~\cite{lag2000}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions} The present study will develop numerical
methods to simulate viscous erosion of multiply-connected
two-dimensional geometries.  By assuming the flow is viscous, we are
assuming the flow is very slow and in small length scales, as is the
case in groundwater flow.  Moreover, by allowing for complex
multiply-connected domains, the effects due to interacting bodies on the
erosion are captured.  We start by authenticating our method on a single
circular body placed symmetrically in a pipe flow.  By artificially
fixing the area of the body as a function of time, we are able to
characterize the long-time behavior of the shape and the resulting flow
and compare these results with analytical results.  Then, we compute the
vorticity of erosion in multiply-connected geometries to better
characterize erosion in porous media.

@@

To further visualize the effect of erosion, we compute the vorticity of
the flow (Figure~\ref{fig:50bodies}) which is equal to the shear stress
in the Stokesian limit.  Finally, to evolve the geometry, a second-order
time stepping method that guarantees that the mesh does not become
distorted will be employed.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations} There are two main limitations of the current
work.  First, the flow is assumed to be two-dimensional.  However, by
assuming that the flow is two-dimensional, we are able to simulate the
erosion of more bodies and with higher resolution than would be
available in three dimensions.  Moreover, Several flows can be described
to good accuracy under a two-dimensional approximation, certain
phenomena occur only in three dimensions.  The boundary integral
equation formulation for the fluid solver exists, and the main
challenges of a BIE solver have been addressed by other groups including
surface representation~\cite{yin-bir-zor2006, vee-rah-bir-zor2011},
near-singular integration~\cite{yin-bir-zor2006, kli-tor2016b}, and fast
summation methods~\cite{yin-bir-zor2004, ros-ols2015}.  The more
significant challenge of extending our work to three dimensions is
guaranteeing that the mesh does not become distorted, which can be done,
for example, by using an extension of the two-dimensional {\thL}
formulation~\cite{amb-sie-tlu2013}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Related work} Simulating eroding bodies has been studied
analytically, experimentally, and numerically.  Because of the scales
involves with geophysical flows, most of these works consider high
Reynolds number flows.  For instance, the experimental work of Moore et
al.~\cite{moo-ris-chi-zha-she2013} considers on the order of $10^4$.
For low Reynolds number flows, Mitchell and Spagnolie~\cite{mit-spa2016}
simulate the erosion of a mobile body in three dimensions, but they only
consider single body.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the paper} In Section~\ref{s:formulation}, we
describe the governing equations for both the fluid and describe the
erosion model using a \thL~framework.  In Section~\ref{s:method}, the
numerical methods in both space and time are described.
Section~\ref{s:results} gives several numerical examples.  Finally,
concluding remarks are made in Section~\ref{s:conclusions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formulation}
\label{s:formulation}
We start by defining the main variables used to model erosion.  We only
consider flows that are confined by a solid wall $\Gamma$ that encloses
$M$ eroding bodies.  The bodies are denoted as $\gamma_\ell$,
$\ell=1,\ldots,M$, and we write $\gamma = \gamma_1 \cup \cdots \cup
\gamma_M$.  Neglecting inertial forces, the dynamics of the fluid is
fully characterized by the position of the bodies $\xx_\ell(s,t) \in
\gamma_\ell$, where $s$ is the arclength and $t$ is time.  Given
$\xx_\ell(s,t)$, $\ell=1,\ldots,M$, derived variables include the fluid
velocity $\uu$, the pressure $p$, and the shear stress $\tau$.   On the
bounding wall, we prescribed a velocity $\UU(\xx,t)$.  Then, the
governing equations are
\begin{equation}
\label{eqn:erosionModel}
\begin{split}
  \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap &&\mbox{conservation
of momentum}\\
\grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap
&&\mbox{conservation of mass} \\
\uu = 0, &\hspace{20pt} \xx \in \gamma, \gap &&\mbox{no slip on the
bodies} \\
\uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap &&\mbox{outer wall
velocity} \\
V_\nn = \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
&&\mbox{erosion model},
\end{split}
\end{equation}
where $V_\nn$ is the normal velocity of the interface and $\mu$ is the
fluid viscosity.  The outer wall is formed by rounding off the corners
of $[-3,3] \times [-1,1]$, and we enforce the Hagen-Poiseuille flow
\begin{align*}
  \uu(\xx) = (1-y^2,0)
\end{align*}
on $\Gamma$.  The boundary of the fluid domain is $\bd\Omega = \gamma_1
\cup \cdots \cup \gamma_M \cup \Gamma$.  To minimize the boundary
effects of the inflow and outflow, we only place bodies in the center
third, $[-1,1] \times [-1,1]$, of the fluid domain.
\begin{figure}[htpb]
  \centering
  \input{figs/schematic.tikz}
  \caption{\label{fig:schematic} A schematic for the governing
    equations.  A no-slip boundary condition is imposed on each body
    $\gamma_\ell$ whose unit normal points outward relative to the
    geometry.  On the outer geometry $\Gamma$, a pipe flow is imposed.
    The bodies are eroded at a rate that is proportional to the
    magnitude of the shear stress applied by the Stokesian fluid.  The
    bodies are constrained to the middle third of the channel that is
    located between the dashed lines.}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods}
\label{s:method}
There are three main numerical methods that need to be developed.  In
Section~\ref{sec:BIE}, we describe how the BIE~\eqref{eqn:completed_BIE}
is solved.  Once the density function is found,
Section~\ref{sec:shearStress} describes how the shear
stress~\eqref{eqn:shearStressLP} is computed.  Then, in
Section~\ref{sec:timeStepping}, a time stepping method for evolving the
interface according to equations~\eqref{eqn:Levo1}
and~\eqref{eqn:thetaevo1} are described.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Post Processing Quantities of Interest}
\label{s:qoi}
To study characteristics of the flow, the vorticity
(Section~\ref{sec:vorticity}), pressure (Section~\ref{sec:pressure}),
and drag (Section~\ref{sec:drag}) will be computed.  Once
equation~\eqref{eqn:completed_DLP} is solved for the density function,
Stokeslets, and rotlets, all of these quantities can be computed.  For
target points in the fluid domain, the resulting layer potentials
involve integrands that are periodic, so the trapezoid rule guarantees
spectral accuracy since the integrand is smooth~\cite{tre-wei2014}.
However, if the target point is close to a boundary, a nearly-singular
integrand must be integrated, and this requires upsampling that is
described in Section~\ref{sec:NSI}.  When the target points is on the
boundary, the integrands involve singularities that are evaluated by
combining singularity subtraction and odd-even
integration~\cite{sid-isr1988}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
\label{s:results}
With the description of the methods complete, we now present numerical results on bodies eroding in Stokes flow. We first discuss the case of a single eroding body, as certain features can be predicted analytically and thus used for validation. We then discuss the erosion of multiple bodies and highlight important differences that arise as compared to the single-body case.
